---
title: "Analysis of the Weight Lifting Excercise Data"
author: "Nobu Asai"
date: "10/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(doParallel)
cl <- makePSOCKcluster(10)
registerDoParallel(cl)
options(warn=-1)
options(scipen=100)
```

#### Executive Summary

This project is to predict the manner of the Unilateral Dumbbell Biceps Curl by using the data captured by Velloso et al ^[1]^. The data were captured by using 4 different devices, "belt", "glove", "arm-band", and "Dumbbel" with accelerometer, gyroscope, and magnetometer. At first, we quickly plotted accelerometer data from "belt", "forearm (glove)", and "arm (arm-band)" againt that from "dumbbel" to see any outstanding correlations.  Then, we tested multiple models to learn with using K-fold cross validation.  By a veficiation with test data, we concluded that Random Forest was the best algorithm to create the model for this data, and observed its accuracy "1" for train-validation data, and "xxx" for test data.


#### 1. About the data

Data was downloaded from the web site below.

```{r echo=TRUE, cache=TRUE}
train_data = read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
test_data = read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
```
There are 19,622 samples with 60 attributes for the train data, and 20 samples with 60 attributes for the test data for predictions.
```{r echo=FALSE, comment=NA}
colnames(train_data)
```

```{r echo=FALSE}
library(gt)

BC_table <- as.data.frame(matrix(ncol=3, nrow=5))
colnames(BC_table) <- c("class","fashion","decision")
BC_table$class <- c('A','B','C','D','E')
BC_table$fashion[1] <- "Exactly according to the specification"
BC_table$fashion[2] <- "Throwing the elbows to the front"
BC_table$fashion[3] <- "Lifting the dumbbell only halfway"
BC_table$fashion[4] <- "Lowering the dumbbell only halfway"
BC_table$fashion[5] <- "Throwing the hips to the front"
BC_table$decision <- "mistake"
BC_table$decision[1] <- "success"

gt(BC_table) %>% 
    tab_options(table.width=pct(90), table.font.size=10, table.align="center")

```



#### Preparing data



```{r echo=TRUE}
picks <- list()
# remove "blank" and "NA" from train/test data
for (i in 1:ncol(train_data)) {
    col_test <- train_data[,c(i)]
    if (sum(is.na(col_test)) == 0) {
        if (sum(col_test == "") == 0) {
            # no blank/NA - to be picked up
            picks <- c(picks, i)
        }
    }
}
train_norm <- train_data[, unlist(picks)]
test_norm <- test_data[, unlist(picks)]

# We do not need "X", "user_name", "raw_timestamp_part_1","raw_timestamp_part_2", "cvtd_timestamp", "new_window", and "num_window" 

train_norm <- train_norm[,8:60]
test_norm <- test_norm[,8:60]
```


```{r echo=TRUE}
# 80% for train and validation, 20% for test
library(caret)
set.seed(154)
train_list <- createDataPartition(train_norm$classe, p=0.8, list=FALSE)
train_cv <- train_norm[train_list,]
test_cv <- train_norm[-train_list,]
```

#### Exploratory analysis

```{r echo=TRUE, warning=FALSE, message=FALSE}
# Setting up libraries for this study

library(ggplot2)
```


```{r echo=FALSE, warning=FALSE, message=FALSE, fig.width=12, fig.height=4}
library(ggplot2)
library(patchwork)

g1 <- ggplot(data=train_norm, aes(x=total_accel_dumbbell, y=total_accel_arm, color=classe))
g1 <- g1 + geom_point()
g1 <- g1 + ggtitle("Acceleration - Dumbbel v.s. Arm")

g2 <- ggplot(data=train_norm, aes(x=total_accel_dumbbell, y=total_accel_belt, color=classe))
g2 <- g2 + geom_point()
g2 <- g2 + ggtitle("Acceleration - Dumbbel v.s. Belt")

g3 <- ggplot(data=train_norm, aes(x=total_accel_dumbbell, y=total_accel_forearm, color=classe))
g3 <- g3 + geom_point()
g3 <- g3 + ggtitle("Acceleration - Dumbbel v.s. Forearm")

g1 | g2 | g3

```



#### Considerations on cross validation

```{r echo=TRUE}
# K-fold cross validation (to minimize overfitting)
train_control <- trainControl(method='repeatedcv', number=10, repeats=10)
```


#### Selecting a model

```{r echo=TRUE, cache=TRUE}
#model_gbm <- train(classe ~ ., data=train_cv, method='gbm', preProcess=c("center","scale"), trControl=train_control, verbose=FALSE)
model_lda <- train(classe ~ ., data=train_cv, method='lda', preProcess=c("center","scale"), trControl=train_control)
model_qda <- train(classe ~ ., data=train_cv, method='qda', preProcess=c("center","scale"), trControl=train_control)
#model_rf <- train(classe ~ ., data=train_cv, method='rf', preProcess=c("center","scale"), trControl=train_control)
```
```{r echo=TRUE}
#gbm_accuracy <- confusionMatrix(train_cv$classe, predict(model_gbm, train_cv))$overall[1]
lda_accuracy <- confusionMatrix(train_cv$classe, predict(model_lda, train_cv))$overall[1]
qda_accuracy <- confusionMatrix(train_cv$classe, predict(model_qda, train_cv))$overall[1]
#rf_accuracy <- confusionMatrix(train_cv$classe, predict(model_rf, train_cv))$overall[1]
```

#####  Evaluation with test data

```{r echo=TRUE}
#confusionMatrix(test_cv$classe, predict(model_gbm, test_cv))$overall[1]
#confusionMatrix(test_cv$classe, predict(model_rf, test_cv))$overall[1]
confusionMatrix(test_cv$classe, predict(model_qda, test_cv))$overall[1]
```


#### Predictions 


```{r echo=TRUE}
#predict(model_gbm, test_norm)
#predict(model_rf, test_norm)
```



##### Reference
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th International Conference in Cooperation with SIGCHI (Augmented Human '13) . Stuttgart, Germany: ACM SIGCHI, 2013.


Ugulino, W.; Cardador, D.; Vega, K.; Velloso, E.; Milidiu, R.; Fuks, H. Wearable Computing: Accelerometers' Data Classification of Body Postures and Movements. Proceedings of 21st Brazilian Symposium on Artificial Intelligence. Advances in Artificial Intelligence - SBIA 2012. In: Lecture Notes in Computer Science. , pp. 52-61. Curitiba, PR: Springer Berlin / Heidelberg, 2012. ISBN 978-3-642-34458-9. DOI: 10.1007/978-3-642-34459-6_6. 





